% Thesis Acknowledgements ------------------------------------------------


%\begin{acknowledgementslong} %uncommenting this line, gives a different acknowledgements heading
%\begin{acknowledgements}      %this creates the heading for the acknowlegments

I would like to acknowledge the following people for their support,
intellectual, moral, emotional, and otherwise.
\begin{itemize}
\item Clay Morrison, my advisor, for his tireless support, for helping
  me stay organized, and for challenging me to be able to explain myself.

\item Kobus Barnard, for turning me on to Bayesian Modeling in the
  first place, way back in the aughts sometime, and for providing the
  resources and opportunities for me to actually learn to put ideas to
  code.

\item The rest of my committee, Kathie Barnes and Helen Zhang, for
  their valuable feedback.

\item Paul Cohen, for bringing me in to SISTA, providing the
  opportunity (and funding!) to work on so many varied challenging 
  research projects, and giving me so much leeway in teaching for the
  information science program.

\item Kyle Simek, Ernesto Brau, and Andrew Prehoehl, for being excellent lab mates,
  for putting together the summer ``coding bootcamps'' that were so
  valuable to me as someone without a lot of formal CS training, for
  the very stimulating reading groups, and for being there to bounce
  ideas off of.

\item My first research student at Oberlin, Bill Huang, for writing code behind
  several of the figures herein.

\item My parents, for being able to smile and nod when I said I was
  going to do {\em another} Ph.D., and for not asking ``have you started
  writing your dissertation yet?'' more than once a week.

\item My children, Arlo and Esai, for being there with snuggles and
  laughter every day, and reminding me what's actually important.

\item My wife, Sarah, for everything and more.  There are no words.
\end{itemize}

%\end{acknowledgements}
%\end{acknowledgmentslong}

% ------------------------------------------------------------------------


